Near Optimal Behavior with Approximate Abstraction

### 2016-02-05 paper/conclusion.tex
adding michael's changes

% Summary of contributions.
Approximate abstraction in \acp{MDP} offers considerable advantages over exact abstraction. In this work we prove bounds for the value lost when behaving according to the optimal policy of the abstract \ac{MDP}. We also empirically demonstrate that approximate abstractions can reduce state space size with negligible loss in the quality of the behavior.

@@

% Future work.
There are many directions for future work.
% Learning abs functions online.
First, we are interested investigating learning various approximate abstraction functions online in the planning or \ac{RL} setting, a strategy first demonstrated by \citeauthor{ortner2013adaptive}.

@@

% Necessary conditions.
Another question of interest is what the necessary conditions are for an abstraction that will preserve any notion of optimality (i.e. the bound between the values under $\pi_G^*$, $\pi_{GA}$ is less than $\textsc{VMax}$). In future work, we hope to identify these conditions.
% Options/termporal abstraction
In the future, we are also interested in characterizing the relationship between temporal abstractions, such as those afforded by Options~\cite{sutton1999between}, and approximate abstractions. Options have been discussed at length in reinforcement learning and planning~\cite{konidaris2007building,chentanez2004intrinsically}. We investigated the use of options in the Taxi domain, and found a negligible difference in the degree and quality of compression compared to compression without Options.
% Compressibility
Lastly, we are interested in understanding the relationship between various approximate abstractions and the information theoretical limitations on the compressibility of certain classes of \acp{MDP}.

>>>>

% Summary of contributions.
Approximate abstraction in \acp{MDP} offers considerable advantages over exact abstraction. In this work, we proved bounds for the value lost when behaving according to the optimal policy of the abstract \ac{MDP}. We also empirically demonstrate that approximate abstractions can reduce state space size with negligible loss in the quality of the behavior.

@@

% Future work.
There are many directions for future work.
% Learning abs functions online.
First, we are interested in investigating learning various approximate abstraction functions online in the planning or \ac{RL} setting, a strategy first demonstrated by \citeauthor{ortner2013adaptive}.
zzz what does that mean?
% Necessary conditions.
Another question of interest is what the necessary conditions are for an abstraction that will preserve any notion of optimality (i.e. the bound between the values under $\pi_G^*$, $\pi_{GA}$ is less than $\textsc{VMax}$).
zzz perhaps too vague
% Options/termporal abstraction
We investigated the use of Options~\cite{sutton1999between} in the Taxi domain, and found a negligible difference in the degree and quality of compression compared to compression without Options. In the future, we are interested in characterizing the relationship between temporal abstractions, such as those afforded by Options, and approximate abstractions.
% Compressibility
Lastly, we are interested in understanding the relationship between various approximate abstractions and the information theoretical limitations on the compressibility of certain classes of \acp{MDP}.

@@

### 2016-02-05 paper/example_domains.tex
adding michael's changes

\section{Example Domains}
To demonstrate the efficacy of these approximate abstraction functions, we test functions in the $\epQ$ abstraction family on four domains -- NChain, Taxi, Minefield and Random. These domains were selected for their variety -- NChain is relatively simple, Taxi is goal-based and hierarchical in nature, Minefield is stochastic, and Random MDP has many near-optimal policies.

% Description of visuals.
We provide graphical visuals of both ground \acp{MDP} and resulting abstract \acp{MDP}. In our visuals, a grey circle indicates a state, while the colored arrows indicate transitions. The thickness of the arrow indicates how much reward is associated with that transition. In the ground \acp{MDP}, states are labeled with a number. In the abstract \acp{MDP} we indicate which ground states were collapsed to each abstract state by labelling the abstract states with their ground states. %(some are visually unwieldy and so were not labelled).

%Description of code base
Our code base (link redacted for review) provides implementations for abstracting arbitrary \acp{MDP} as well as visualizing and evaluating the resulting abstract \acp{MDP}. We use the graph visualization library GraphStream~\cite{graphstream} and planning and \ac{RL} library, BURLAP~\cite{BURLAPCITATION}.

@@

NChain, pictured in Figure~\ref{fig:nchain-visual}, is a simple MDP heavily investigated in the Bayesian Reinforcement Learning literature due to the interesting exploration problem it poses~\cite{dearden1998bayesian}. In all states, the agent has two actions available: advance down the chain, or return to state 0. The agent receives a small positive reward, $r_{small}$ for returning to state 0, and no reward for advancing down the chain, unless the agent transitions to the last state in the chain which provides a large positive reward, $r_{large}$. Transitions also have small slip probability, $\rho$, such that the applied action results in the wrong action. In our implementation, we set $N=10$, $\rho=0.2$, $r_{small} = 0.2$, and $r_{large} = 1.0$.

@@

Taxi has long been studied by the hierarchical reinforcement learning literature~\cite{dietterich2000hierarchical}. The agent, operating in a grid world style domain may move left, right, up, and down, as well as pickup a passenger and drop off a passenger. The goal is achieved when the agent has taken all passenger to their destinations.

@@

Minefield is a particular variant of Grid World \cite{russell1995modern} with slip probability of $x$. The reward function is such that moving up in the top row of the grid receive $1.0$ reward, all other transitions receive $0.2$ reward, except for transitions to a random set of $\kappa$ states (which may include the top row) that receive $0$ reward (these are the states with mines in them, thus the name). During experimentation, we set $N=10, M=4, \epsilon=1.1, \kappa = 5, x = 0.01$. We visualize the original Minefield and the abstract Minefield in Figure~\ref{fig:minefield-visual}.

% Minefield Compression Visuals
\begin{figure}
\subfigure[Ground Minefield]{
\includegraphics[width=0.48\columnwidth]{figures/ground_minefield.png}}
\subfigure[Abstract Minefield]{
\includegraphics[width=0.48\columnwidth]{figures/abs_minefield.png}}
\label{fig:minefield-visual}
\caption{Comparison of ground Minefield \ac{MDP} and abstract \ac{MDP}, under a $\epQ$ abstraction, with $\epsilon=1.1$}
\end{figure}

@@

The Random \ac{MDP} domain we consider is as follows. There are $N$ states, and $K$ actions. For each state, each action can transition to two randomly selected (but fixed) states, each with probability $0.5$.

>>>>

\section{Example Domains}
To illustrate our approximate abstraction functions, we apply $\epQ$ abstraction to four example domains---NChain, Taxi, Minefield and Random. These domains were selected for their diversity---NChain is relatively simple, Taxi is goal-based and hierarchical in nature, Minefield is stochastic, and Random MDP has many near-optimal policies.

% Description of visuals.
We provide visuals of both ground \acp{MDP} and resulting abstract \acp{MDP} in which a grey circle indicates a state and colored arrows indicate transitions. The thickness of the arrow indicates how much reward is associated with that transition. In the ground \acp{MDP}, states are labeled with a number. In the abstract \acp{MDP}, we indicate which ground states were collapsed to each abstract state by labelling the abstract states with their ground states. %(some are visually unwieldy and so were not labelled).

%Description of code base
Our code base (link redacted for review) provides implementations for abstracting arbitrary \acp{MDP} as well as visualizing and evaluating the resulting abstract \acp{MDP}. We use the graph-visualization library GraphStream~\cite{graphstream} and the planning and \ac{RL} library, BURLAP~\cite{BURLAPCITATION}.

@@

NChain, pictured in Figure~\ref{fig:nchain-visual}, is a simple MDP heavily investigated in the Bayesian RL literature due to the interesting exploration problem it poses~\cite{dearden1998bayesian}. In all states, the agent has two actions available: advance down the chain, or return to state 0. The agent receives a small positive reward, $r_{small}$ for returning to state 0, and no reward for advancing down the chain. The exception is that when the agent transitions to the last state in the chain, it receives a large positive reward, $r_{large}$. Transitions also have small slip probability $\rho$, such that the applied action results in the opposite dynamics. In our implementation, we set $N=10$, $\rho=0.2$, $r_{small} = 0.2$, and $r_{large} = 1.0$.

@@

Taxi has long been studied by the hierarchical RL literature~\cite{dietterich2000hierarchical}. The agent, operating in a grid-world style domain, may move left, right, up, and down, as well as pick up a passenger and drop off a passenger. The goal is achieved when the agent has taken all passengers to their destinations.

@@

Minefield is a test problem we are introducing that uses the Grid World dynamics of \citeauthor{russell1995modern} with slip probability of $x$. The reward function is such that moving up in the top row of the grid receives $1.0$ reward; all other transitions receive $0.2$ reward, except for transitions to a random set of $\kappa$ states (which may include the top row) that receive $0$ reward. (These are the states with ``mines'' in them.) We set $N=10, M=4, \epsilon=1.1, \kappa = 5, x = 0.01$. We depict the original Minefield and an abstract Minefield in Figure~\ref{fig:minefield-visual}.

zzz problem: text says ``top row'', but image is oriented diagonally.

% Minefield Compression Visuals
\begin{figure}
\subfigure[Ground Minefield]{
\includegraphics[width=0.48\columnwidth]{figures/ground_minefield.png}}
\subfigure[Abstract Minefield]{
\includegraphics[width=0.48\columnwidth]{figures/abs_minefield.png}}
\label{fig:minefield-visual}
\caption{Comparison of ground Minefield \ac{MDP} and abstract \ac{MDP}, under a $\epQ$ abstraction with $\epsilon=1.1$}
\end{figure} 
zzz the lines are very thin... not visible to me.

@@

In the Random \ac{MDP} domain we consider, there are $N$ states and $K$ actions. For each state, each action transitions to one of two randomly selected states, each with probability $0.5$.

### 2016-02-05 paper/results_discussion.tex
adding michael's changes

\section{Approximate Abstraction on Example Domains}
%Explanation of what abstraction is used
We run experiments on aggregation functions in the $\epQ$ family of approximate abstractions. For each trial our aggregation greedily groups ground states into abstract states that in a way that satisfies the $\epQ$ criteria. The order in which ground states are grouped is randomized, so there is randomization across trials. Every ground state is equally weighted in its abstract state.

%Explanation of what plots done
For each domain we report two quantities as a function of epsilon with 95\% confidence bars. First, we compare the number of states in the abstract \ac{MDP} for different values of $\epsilon$, shown in Figure~\ref{fig:main_empirical_results1}. The smaller the number of abstract states, the smaller the state space of the $\ac{MDP}$ that the agent must plan over. Second, we report the value under the abstract policy of the initial ground state, shown in Figure~\ref{fig:main_empirical_results2}. In the Taxi and Random domains, 200 trials are run for each data point, whereas 20 trials are run for data points in Minefield and NChain.

%Discuss results
Our empirical results corroborate our thesis -- that approximate state abstractions can form the basis useful state aggregation criteria. In both NChain and Minefield, we observe that as $\epsilon$ increases from $0$, the number of states which must be planned over is reduced, and optimal behavior is either fully maintained (NChain) or very nearly maintained (Minefield). Similarly for Taxi, when $\epsilon$ is between $.02$ and $.025$ we observe a reduction in the number of states in the abstract \ac{MDP} while value is fully maintained. After $.025$, increased reduction in state space size comes at a cost of value. Lastly, as $\epsilon$ is increased in the Random domain, there is a smooth reduction in the number of abstract states with a corresponding cost in the value of the derived policy.

Our experimental results also highlight a noteworthy characteristic of approximate state abstraction in goal-based \acp{MDP}. Taxi exhibits relative stability in state space size and behavior for epsilon up to $.02$ until both fall off dramatically. We attribute the sudden fall off of these quantities to its goal-based nature; once information critical for achieving optimal behavior is lost in the state aggregation, solving the goal -- and so acquiring any reward -- is unattainable. Conversely, in the Random domain, a great deal of near-optimal policies are available to the agent. Thus, even the information for optimal behavior is lost with an increase in $\epsilon$ there are a great deal of near-optimal policies available to the agent which are not lost by the abstraction.

>>>>

\section{Approximate Abstraction on Example Domains}
%Explanation of what abstraction is used
We ran experiments on aggregation functions in the $\epQ$ family of approximate abstractions. 
zzz why? wouldn't it be better to show all 4?
For the purpose of illustrating what kinds of approximations are possible, we built each abstraction by first solving the MDP, then greedily aggregating ground states into abstract states that satisfied the $\epQ$ criteria. Since this approach represents an order-dependent approximation to the maximum amount of abstraction possible, we randomized the order in which states were considered across trials. Every ground state is equally weighted in its abstract state.

%Explanation of what plots done
For each domain, we report two quantities as a function of epsilon with 95\% confidence bars. First, we compare the number of states in the abstract \ac{MDP} for different values of $\epsilon$, shown in Figure~\ref{fig:main_empirical_results1}. The smaller the number of abstract states, the smaller the state space of the $\ac{MDP}$ that the agent must plan over. Second, we report the value under the abstract policy of the initial ground state, shown in Figure~\ref{fig:main_empirical_results2} so the tradeoff between abstract state space size and near optimality is made evident. In the Taxi and Random domains, 200 trials were run for each data point, whereas 20 trials were sufficient for data points in Minefield and NChain.

zzz the font size in the figures needs to be larger---too tiny for my eyes

%Discuss results
Our empirical results corroborate our thesis---approximate state abstractions can decrease state space size while retaining bounded error. In both NChain and Minefield, we observe that, as $\epsilon$ increases from $0$, the number of states that must be planned over is reduced, and optimal behavior is either fully maintained (NChain) or very nearly maintained (Minefield). Similarly for Taxi, when $\epsilon$ is between $.02$ and $.025$, we observe a reduction in the number of states in the abstract \ac{MDP} while value is fully maintained. After $.025$, increased reduction in state space size comes at a cost of value. Lastly, as $\epsilon$ is increased in the Random domain, there is a smooth reduction in the number of abstract states with a corresponding cost in the value of the derived policy.

zzz why not include all the graphs in one figure?

zzz wait, why do the policies in the highly abstracted random MDPs end up statistically significantly worse than random?? I'm not convinced these error bars are correct.

Our experimental results also highlight a noteworthy characteristic of approximate state abstraction in goal-based \acp{MDP}. Taxi exhibits relative stability in state space size and behavior for $\epsilon$ up to $.02$, at which point both fall off dramatically. We attribute the sudden fall off of these quantities to the goal-based nature of the domain; once information critical for achieving optimal behavior is lost in the state aggregation, solving the goal---and so acquiring any reward---is impossible. Conversely, in the Random domain, a great deal of near-optimal policies are available to the agent. Thus, even as the information for optimal behavior is lost with increasng $\epsilon$, there are many near-optimal policies available to the agent that remain available.
